### 2024-04-02 paper/paper.qmd
update abstract and introduction

---
title: "Degrees of Difference: Analyzing the Impact of Education on Earnings in Canada"
author: 
  - Yuanyi (Leo) Liu
thanks: "Code and data are available at: https://github.com/leoyliu/Analyzing-the-Impact-of-Education-on-Earnings-in-Canada"
date: today
date-format: long
abstract: "In this paper, we explore the connection between education and hourly wages in Canada, focusing on data from the year 2000. Our findings indicate a clear trend: higher educational levels correlate with increased average hourly wages across all age categories. This research highlights the importance of education in determining earning potential and suggests that investment in education could have long-term economic benefits. The study provides evidence for policy implications regarding educational incentives and workforce development strategies in Canada."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

# install.packages("tidyverse")
# install.packages("dplyr")
# install.packages("here")
# install.packages("mapdata")
# install.packages("ggplot2")
# install.packages("haven")
# install.packages("kableExtra")

library(tidyverse)
library(dplyr)
library(here)
library(mapdata)
library(ggplot2)
library(haven)
library(kableExtra)
```

# Introduction {#sec-intro}

The relationship between education and earnings is a well-established topic of interest within the field of labor economics, providing insights into the broader socio-economic fabric of a country. As Canada's economy continues to diversify and specialize in various sectors, the value of education in this landscape remains a critical question for policymakers and the public. While previous studies have explored this link, there is an ongoing need to update and deepen our understanding of how this dynamic plays out across different age groups in the modern economy.

This paper aims to fill the knowledge gap by analyzing the impact of education levels on the average hourly wage rate in Canada, with a specific focus on three age cohorts: young entrants (15-24 years), established workers (25-54 years), and senior professionals (55 years and over). The analysis begins in the year 2000, a period of significant technological advancement and economic change, offering a contemporary perspective on the education-wage relationship. By dissecting the data across these age groups, the study provides nuanced insights into the economic returns of education at various career stages.

Our research methodology involved a detailed statistical analysis of wage data, stratified by education level and age group, to uncover patterns and variations in earnings. The findings confirm the hypothesis that higher education correlates with increased wages, with the effect becoming more significant as workers age. This not only reinforces the case for educational investment but also suggests that such investment pays dividends throughout an individual’s career, not just in the early years.

In this study, our primary focus is to estimate the effect of education on hourly wages across different age groups in Canada. Our estimand is the incremental average hourly wage rate for individuals within the specified education levels, compared to the next lower education level, while controlling for age. By quantifying this effect, we aim to capture the economic value of educational attainment and how it translates into wage premiums across various stages of a typical Canadian worker’s career span.

The structure of the paper is organized as follows: Following @sec-intro, @sec-data presents the data, detailing the data sources, analytical techniques, and the rationale behind the chosen methods. @sec-model then delves into the specifics of the linear model analysis, laying out the statistical underpinnings that support our investigation. After that, @sec-result discusses the results, elaborating on the observed trends and patterns in wage rate data. @sec-discussion interprets these findings in light of the current economic and educational context in Canada, exploring potential factors influencing these trends, drawing connections to broader socio-economic issues, and providing suggestions for future research in this area.

# Data {#sec-data}

This section aims to offer an insightful understanding of the dataset utilized in our analysis. The dataset captures the birth rates across various demographics in the United States from 1980 to 2020. The data provide a broader context, allowing an analysis of trends over time and across different economic cycles, including the period of the Great Recession.

## Source and Methodology
Data on birth rates for women aged 15-44 were sourced from National Vital Statistics Reports covering 2015, 2019, and 2020 [@cite2015birth; @cite2019birth; @cite2020birth], compiled from birth certificates across the US and the District of Columbia. @citeoriginal provided aggregated birth data for six cohorts over five-year spans, derived from the NBER Natality Database [@citeNBER] and NCHS microdata [@citeNCHS] for 1980-2019. In this study, we simply used the replication data package provided by @citeoriginal for analysis.

While there were alternative datasets available from other public and private sources, this dataset was chosen due to its comprehensive coverage, reliability, and the level of detail it offers. It includes data from the National Center for Health Statistics [@citeNCHS] and the U.S. Census Bureau, which are both reputable sources of demographic data.

The data was processed and cleaned using R[@citeR], a powerful statistical programming language. For key operations, please refer to the [@sec-appendix].

## Variables

To better understand the data, a summary table was developed to provide a detailed description of each variable, explaining its relevance and how it contributes to our understanding of the topic. Our focus is not only on the direct measures of birth rates but also on a range of demographic and economic variables that offer a comprehensive view of the factors influencing fertility decisions during the recessionary period.

## Measurements

The measurement of birth rates in our study involves a detailed analysis of the annual number of live births per 1,000 women in various age groups, ranging from 15 to 44 years. These rates were extracted from datasets provided by the National Center for Health Statistics (NCHS) [@citeNCHS], ensuring accuracy and reliability in capturing demographic trends across the United States from 1980 to 2020.

For each age group, birth rates were calculated by dividing the total number of live births by the population of women in that age range, then multiplying by 1,000 to standardize the measure. This approach allows for comparison across different demographics and time periods, providing a clear picture of how birth rates have shifted, particularly in response to economic conditions such as the Great Recession.

In addition to age-specific birth rates, state-level data were also analyzed to identify geographic variations in birth rate trends. This involved mapping birth rates against state populations, taking into consideration the exclusion of data from states like Alaska and Hawaii where mapping constraints exist. Further details on this aspect of our study are discussed in @sec-result.

# Model {#sec-model}

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results {#sec-result}

@sec-result presents the core findings from our analysis of US birth rates, specifically focusing on the changes observed over time, across different age groups of young women, and by state.

In conclusion, our findings provide a complete perspective of dropping birth rates in the United States, demonstrating both temporal and demographic tendencies. The upcoming @sec-discussion will look at the various variables and biases that may be driving these trends.

\newpage

# Discussion {#sec-discussion}

The findings of this study offer a layered perspective on the decline in birth rates among young people in the United States, particularly in the context of the Great Recession. Our analysis provides an understanding of the multifaceted influences on demographic trends.

## Findings

In our replication of @citeoriginal, we reaffirmed the central finding that the Great Recession has had a lasting impact on birth rates among young people in the United States. Our analysis of Figures 1 and 2 showed a pronounced decline in birth rates across all age groups, with the steepest decrease observed among those aged 15-29. This trend extended beyond the economic recovery, suggesting that the Great Recession may have fundamentally altered the family planning trajectory of a generation.

## Economic Impact Insights

The dramatic drop in birth rates among young Americans during and after the Great Recession demonstrates the profound impact of economic hardship on reproductive decisions. This decline in birth rates is not simply coincidental with the date of the recession, but rather indicates a deeper, more systemic influence of economic insecurity on personal life choices, notably the decision to establish or expand a family. The persistence of this pattern, even in the years after economic recovery, suggests that the effects of the recession went beyond immediate financial hardship, influencing long-term views of financial security and stability. This is further evidenced by the lack of rebound in birth rates post-recession, which might have been anticipated if the decline were solely due to immediate economic pressures.

The economic model of fertility provides a framework for understanding these patterns by taking into account the cost of pregnancy and raising as well as the opportunity costs associated with parental time and resources. The recession likely increased these expenditures and opportunity costs, making the decision to have children more difficult. Unemployment and job uncertainty may have caused a reevaluation of the feasibility of having children, resulting in the postponement or avoidance of childbirth. Furthermore, the economic downturn may have shifted young adults' expectations and desires for financial security, impacting their family planning decisions.

## Societal and Technological Influences

Apart from economic factors, the decline in birth rates among young people also reflects broader societal and technological shifts. The period following the Great Recession coincided with significant changes in social norms, increased educational and career opportunities for women, and advancements in reproductive technology. These factors collectively have empowered individuals, especially women, to make more independent decisions regarding their reproductive health and family planning.

As women's educational attainment and labor force involvement have increased, so has the opportunity cost of childbirth, potentially contributing to the drop in birth rates. Technological developments have given women greater control over their fertility, allowing for more intentional planning around childbirth. Social changes, such as delayed marriage and a growing acceptance of childlessness or smaller family sizes as realistic lifestyle options, further compound these tendencies.

The interaction of these economic, social, and technical elements has altered the landscape of family planning for young Americans. While the economic model of fertility provides a framework for understanding these developments, including sociological and technical effects provides a more complete view of the dynamics affecting current birth rate patterns.

## Weaknesses and Future Research Directions

One limitation of our study is the potential for unobserved variables that could affect birth rates, such as cultural shifts and changes in social norms, which were not fully captured in the data. Additionally, the original study did not account for the influence of the Affordable Care Act and its potential impact on family planning decisions, an area our study also does not explore.

Future research should aim to disentangle these complex relationships further, perhaps through longitudinal studies or by incorporating more detailed data on individual socioeconomic status. Understanding these dynamics is crucial for developing policies that support young people in their family planning decisions during and after economic downturns.

\newpage

\appendix

# Appendix {-}

# Data Manipulation and Cleaning {#sec-appendix}

Most of the data in our dataset was previously cleaned for the project we are replicating. Thus, @fig-trends-in-birth-rate directly used data from `outputs/data/fig_1.csv` without any further cleaning necessary, as it simply displays the birth rate from 1980 to 2020. Thus, the cleaned data is directly stored to `outputs/data/fig_1.csv` from `inputs/data/fig_1.csv`.

In @fig-trends-in-birth-rate-in-young-people, the dataset was imported from `inputs/data/fig_2.csv` and used `dpylr` [@citedplyr] to select the birth rates of the six age groups (from 15 to 44). Furthermore, the selected data was renamed using `tidyverse` [@citetidyverse] and changed into names that represent the content of the data better. Moreover, these data is then pivoted using `tidyverse` [@citetidyverse] in order to be graphed properly using `ggplot2` [@citeggplot].

In @fig-birth-rate-comparison-old-and-young, the first dataset was imported from `inputs/data/fig_3.csv` and the dataset containing American states information was from `mapdata` [@citemapdata]. The first dataset was first sliced to remove birth rate information regarding the states of Alaska and Hawaii, as these two states can't be properly shown in the dataset from `mapdata` [@citemapdata]. Then, a temporary data frame is created with tibble from `tidyverse` [@citetidyverse] with the abbreviated state names from `inputs/data/fig_3.csv` and the state full names. Moreover, the abbreviated state name in the first dataset is switched with the state full name using left_join from `dpylr` [@citedplyr], which is then selected and renamed using `tidyverse` [@citetidyverse]. After this, the adjusted dataset with the state full names and birth rate is merged with the second dataset that contains the American states information for further graphing with `haven` [@citehaven].

In all three figures, the library `here` [@citehere] was used to ensure that the file path should be accessible in all directories.

\newpage

# References

### 2024-04-19 paper/paper.qmd
updated paper structure and scripts, added sketches

@@
 abstract: "In this paper, we explore the connection between education and hourly wages in Canada, focusing on data from the year 2000.
-Our findings indicate a clear trend: higher educational levels correlate with increased average hourly wages across all age categories.
+Our findings indicate a clear trend: higher educational levels correlate with increased average hourly wages for individuals aged 25 to 54.
 This research highlights the importance of education in determining earning potential and suggests that investment in education could have long-term economic benefits.
@@
 # install.packages("kableExtra")
+# install.packages('tinytex')
+# tinytex::install_tinytex()
+# to uninstall TinyTeX, run tinytex::uninstall_tinytex()
 
@@
 library(kableExtra)
+library(tinytex)
 ```
@@
 As Canada's economy continues to diversify and specialize in various sectors, the value of education in this landscape remains a critical question for policymakers and the public.
-While previous studies have explored this link, there is an ongoing need to update and deepen our understanding of how this dynamic plays out across different age groups in the modern economy.
+While previous studies have explored this link, there is an ongoing need to update and deepen our understanding of how this dynamic plays out in the modern economy.
 
-This paper aims to fill the knowledge gap by analyzing the impact of education levels on the average hourly wage rate in Canada, with a specific focus on three age cohorts: young entrants (15-24 years), established workers (25-54 years), and senior professionals (55 years and over).
-The analysis begins in the year 2000, a period of significant technological advancement and economic change, offering a contemporary perspective on the education-wage relationship.
-By dissecting the data across these age groups, the study provides nuanced insights into the economic returns of education at various career stages.
+This paper narrows the focus to the Canadian labor market, where the interplay between educational achievement and wages within the core working age group, those aged 25-54, offers a mirror to the efficacy and value of educational advancements in the workplace.
+This demographic is selected for its relevance to the labor force and policy implications, excluding younger individuals still likely in the education system and older individuals, who may present outlier educational experiences that do not align with the central workforce.
 
-Our research methodology involved a detailed statistical analysis of wage data, stratified by education level and age group, to uncover patterns and variations in earnings.
-The findings confirm the hypothesis that higher education correlates with increased wages, with the effect becoming more significant as workers age.
-This not only reinforces the case for educational investment but also suggests that such investment pays dividends throughout an individual’s career, not just in the early years.
+This investigation seeks to address a gap in the current literature by providing a targeted analysis of the 25-54 age group in Canada since the year 2000, a period marked by rapid technological change and economic evolution.
+By applying a linear model to wage data categorized by education level, this study uncovers a positive correlation between educational attainment and hourly wages, suggesting that higher education can be linked to improved wage outcomes in this key demographic.
 
-In this study, our primary focus is to estimate the effect of education on hourly wages across different age groups in Canada.
-Our estimand is the incremental average hourly wage rate for individuals within the specified education levels, compared to the next lower education level, while controlling for age.
+In this study, our primary focus is to estimate the effect of education on hourly wages in Canada.
+Our estimand is the incremental average hourly wage rate for individuals within the specified education levels, compared to the next lower education level.
 By quantifying this effect, we aim to capture the economic value of educational attainment and how it translates into wage premiums across various stages of a typical Canadian worker’s career span.
@@
 
-This section aims to offer an insightful understanding of the dataset utilized in our analysis.
-The dataset captures the birth rates across various demographics in the United States from 1980 to 2020.
-The data provide a broader context, allowing an analysis of trends over time and across different economic cycles, including the period of the Great Recession.
+This section aims to offer an insightful understanding of the dataset utilized in our analysis, which serves as the foundation for our examination of the relationship between education and average hourly wage rates in Canada, specifically within the 25-54 age demographic.
 
 ## Source and Methodology
-Data on birth rates for women aged 15-44 were sourced from National Vital Statistics Reports covering 2015, 2019, and 2020 [@cite2015birth; @cite2019birth; @cite2020birth], compiled from birth certificates across the US and the District of Columbia.
-@citeoriginal provided aggregated birth data for six cohorts over five-year spans, derived from the NBER Natality Database [@citeNBER] and NCHS microdata [@citeNCHS] for 1980-2019.
-In this study, we simply used the replication data package provided by @citeoriginal for analysis.
 
-While there were alternative datasets available from other public and private sources, this dataset was chosen due to its comprehensive coverage, reliability, and the level of detail it offers.
-It includes data from the National Center for Health Statistics [@citeNCHS] and the U.S.
-Census Bureau, which are both reputable sources of demographic data.
+This study utilizes a dataset sourced from the Open Government Portal of Canada [], specifically designed to track the correlation between educational attainment and hourly wages across various demographic segments of the Canadian workforce.
+The dataset covers the period from the year 2000 onwards, providing a longitudinal view of wage trends in relation to educational background.
 
-The data was processed and cleaned using R[@citeR], a powerful statistical programming language. For key operations, please refer to the [@sec-appendix].
+Alternative datasets, such as those from Statistics Canada’s Labour Force Survey, were considered but were not selected due to their less detailed categorization of education levels and their broader focus on employment without specific wage breakdowns by education.
+
+The data was processed and cleaned using R[@citeR], a powerful statistical programming language.
+The selection of this dataset was motivated by its direct relevance to the research question and its regular updates, ensuring that the data remains relevant for observing current trends.
+Initial data processing involved filtering the dataset to focus solely on entries categorized under 'Canada' in geography, eliminating data irrelevant to the national focus of this research.
+This step was crucial to maintaining the clarity and relevance of the analysis.
+For key operations, please refer to the [@sec-appendix].
 
@@
 
-To better understand the data, a summary table was developed to provide a detailed description of each variable, explaining its relevance and how it contributes to our understanding of the topic.
-Our focus is not only on the direct measures of birth rates but also on a range of demographic and economic variables that offer a comprehensive view of the factors influencing fertility decisions during the recessionary period.
+To better understand data, key variables extracted for this study include the year of data collection, education level, age group, and the average hourly wage rate.
+Notably, we have streamlined the 'Education level' from a broad range of categories into a numeric variable that aligns with ascending educational attainment, facilitating a quantitative analysis of its impact on wages.
+The age group has been limited to the 25-54 years range to focus on the most economically active segment of the population, avoiding potential outliers from younger individuals with less work experience and older individuals whose advanced education, such as doctoral degrees, could distort the analysis.
+Wage rates are presented in Canadian dollars per hour, reflecting pre-tax earnings, inclusive of tips, commissions, and bonuses.
+
+```{r}
+#| label: tbl-display-clean-data
+#| tbl-cap: First Ten Rows of Wage Rates Spanning from 2000 to 2019 For Cleaned Data
+#| echo: false
+
+cleaned_data <- read.csv(here::here("data/analysis_data/analysis_data.csv"))
+
+cleaned_data |>
+  select(Year, Education.level, Age.group, Avg.hourly.wage.rate) |>
+  slice(1:10) |>
+  kable(
+    col.names = c("Year", "Education Level", "Age", "Hourly Wage"),
+    digits = 1,
+    booktabs = TRUE
+  )
+```
+@tbl-display-clean-data, created with `kableExtra` [@citekableExtra], outlines the first ten records of our dataset, featuring average hourly wages for Canadians aged 25-54 across varying levels of education from the years 2000 to 2019.
+This snapshot reveals the variable structure of our analysis, which includes the `Year` of wage data, the defined `Education level` ranging from "0 - 8 years" to "Above bachelor’s degree," and the consistent `Age group` focus.
+It provides an early indication of the ascending trend in wages with higher educational attainment, setting the stage for our deeper investigation into the economic value of education within Canada's workforce.
+
+To transition from a tabular overview to a more graphical representation, we will visualize the data, which will allow us to spot patterns and trends that are not immediately apparent in numerical form.
+
+```{r fig.width=7, fig.height=3}
+#| label: fig-display-clean-data
+#| fig-cap: Visualization of Wage Rate from 2000 to 2019
+#| echo: false
+
+ggplot(cleaned_data, aes(x = Year, y = Avg.hourly.wage.rate, color = Education.level)) +
+  geom_point(alpha = 0.7, size = 2) + 
+  labs(
+    x = "Year",
+    y = "Average Hourly Wage Rate (CAD)",
+    color = "Education Level"
+  ) +
+  theme_minimal(base_size = 10) +  # Reduced base_size for smaller overall text
+  theme(
+    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8),  # Smaller font for x-axis labels
+    axis.text.y = element_text(size = 8),  # Smaller font for y-axis labels
+    legend.position = "right",
+    legend.text = element_text(size = 8)  # Smaller font for legend text
+  ) +
+  scale_color_brewer(palette = "Set1")  # Keep the color palette for distinction
+```
+@fig-display-clean-data illustrates the distribution of average hourly wage rates across different educational levels from 2000 to 2019 for Canadian workers aged 25-54.
+Each dot represents the wage rate for a specific education level in a given year, with color coding to distinguish between educational categories, ranging from "0 – 8 years" to "University degree." The spread and upward trend of the points suggest that individuals with higher educational qualifications tend to have higher average hourly wages, and this trend persists over the 20-year period.
 
@@
 
-The measurement of birth rates in our study involves a detailed analysis of the annual number of live births per 1,000 women in various age groups, ranging from 15 to 44 years.
-These rates were extracted from datasets provided by the National Center for Health Statistics (NCHS) [@citeNCHS], ensuring accuracy and reliability in capturing demographic trends across the United States from 1980 to 2020.
+The measurement of variables within this dataset was handled to ensure accuracy and relevance.
+The 'Average Hourly Wage Rate' is calculated based on actual wages reported by employers, providing an objective measure of income.
+This data is then vetted and standardized by statistical agencies to ensure it reflects accurate and fair representations of wages across different demographics.
 
-For each age group, birth rates were calculated by dividing the total number of live births by the population of women in that age range, then multiplying by 1,000 to standardize the measure.
-This approach allows for comparison across different demographics and time periods, providing a clear picture of how birth rates have shifted, particularly in response to economic conditions such as the Great Recession.
+For educational attainment, data entry occurs as individuals enter the workforce or update their qualifications, with categorizations reflecting the highest level of formal education completed.
+This data is collected through surveys and employment records, often as part of broader demographic data collection efforts by governmental agencies.
 
-In addition to age-specific birth rates, state-level data were also analyzed to identify geographic variations in birth rate trends.
-This involved mapping birth rates against state populations, taking into consideration the exclusion of data from states like Alaska and Hawaii where mapping constraints exist.
-Further details on this aspect of our study are discussed in @sec-result.
+For an in-depth technical breakdown of the statistical analyses and additional visualizations, readers are directed to the appendices, where these aspects are discussed in detail.
+This ensures that the main text remains accessible while still providing resources for more technically inclined readers.
+
+Correlations between the variables will be examined at @sec-model, providing insights into how education levels are associated with hourly wages.
+By understanding these relationships, we can better infer the potential impact of education on earnings within the Canadian economy.
 
@@
 
-The goal of our modelling strategy is twofold. Firstly,...
+The objective of our modeling approach is to quantify the relationship between educational attainment and hourly wages.
+Our analysis employs a Bayesian framework to assess how changes in the level of education correlate with variations in wage rates among Canadian workers aged 25-54.
 
@@
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+Define $y_i$ represent the average hourly wage rate for the $i^{th}$ individual.
+The predictor variable, $x_i$, corresponds to the numeric value assigned to each education level.
+The model can be described by the following equations:
 
-\begin{align} 
+$$
+\begin{aligned}
 y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
+\mu_i &= \beta_0 +\beta_1x_i\\
+\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
+\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
+\sigma &\sim \mbox{Exponential}(1) \\
+\end{aligned}
+$$
 We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
 
-### Model justification
+The model is executed in R[@citeR] using the rstanarm package [@rstanarm].
+Default priors from rstanarm are utilized, set with a mean of zero and a conservative standard deviation, reflecting no strong prior beliefs about the magnitude of the education coefficients.
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+## Model justification
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
+Given the economic theory and previous empirical findings, we anticipate that higher educational qualifications would correspond to higher hourly wages.
+This is encapsulated in the assumption that $\beta_i$ should have a positive effect on $y_i$.
+Through Bayesian analysis, we aim to capture the degree to which educational attainment can predict wage rates, thus providing empirical evidence to support or refine this hypothesis.
+
+Mathematical notation is incorporated using LaTeX syntax, enabling precise formulation of our statistical model. For example, the average effect of education on hourly wage rate is denoted by $\beta$.
 
@@
 
+The transformation from raw to analysis-ready data involved substantial cleaning, including filtering by geography (Canada), type of work (both full- and part-time), and specific wage type (average hourly rate).
+Moreover, we excluded broad or ambiguous categories like 'Some high school' and 'Total, all education levels' to ensure clarity in our analysis.
+The resulting 'Education_numeric' variable was constructed using a sorted factorization process, mapping each education level to a corresponding numeric value.
+
+In presenting our findings, we will incorporate graphs that visualize the relationship between education and wage rates, providing an intuitive understanding of the data's characteristics.
+Tables summarizing the observations, along with discussions on the graphs, will be included to elucidate key trends and observations.
+Summary statistics and any notable relationships between the variables will also be featured, offering a detailed overview of the dataset.
+Should the volume of this information warrant it additional detail, these elements may be relegated to appendices to maintain the accessibility of the data section.
+
 \newpage

### 2024-04-19 scripts/01-simulate_data.R

@@ -20,28 +20,45 @@ library(dplyr)
 
 set.seed(853)
 
-# Initialize a tibble
-simulated_data <- tibble(
-  year = c(1980:2020),
-  br_1519 = abs(rnorm(n = 41,mean = 50.5, sd=10.5)),
-  br_2024 = abs(rnorm(n = 41,mean = 55.5, sd=10.5)),
-  br_2529 = abs(rnorm(n = 41,mean = 80.5, sd=5.5)),
-  br_3034 = abs(rnorm(n = 41,mean = 55.5, sd=5.5)),
-  br_3539 = abs(rnorm(n = 41,mean = 35.5, sd=5.5)),
-  br_4044 = abs(rnorm(n = 41,mean = 25.5, sd=5.5)),
-)
+# Set parameters for simulation
+years <- 2000:2019
+education_levels <- c("0 - 8 years", "High school graduate", 
+                      "Post-secondary certificate or diploma", "Trade certificate or diploma", 
+                      "Community college, CEGEP", "University certificate below bachelors degree", 
+                      "University degree", "Bachelor's degree", "Above bachelor's degree")
+age_group <- "25-54 years"
 
-simulated_data$Younger <- simulated_data |>
-  select(
-    br_1519,br_2024,br_2529
-  ) |>
-  rowMeans()
+# Simulate data
+simulated_data <- expand.grid(Year = years, 
+                              Education_level = education_levels, 
+                              Age_group = age_group)
 
-simulated_data$Older <- simulated_data |>
-  select(
-    br_3034,br_3539,br_4044
-  ) |>
-  rowMeans()
+# Assign education_numeric based on order in education_levels
+simulated_data$Education_numeric <- as.integer(factor(simulated_data$Education_level, levels = education_levels))
 
-# View the simulated data
-print(simulated_data)
+# Define the wage rate ranges for each education level
+wage_rate_ranges <- list(
+  `0 - 8 years` = c(11, 13),
+  `High school graduate` = c(12, 14),
+  `Post-secondary certificate or diploma` = c(13, 15),
+  `Trade certificate or diploma` = c(14, 16),  
+  `Community college, CEGEP` = c(16, 18),
+  `University certificate below bachelors degree` = c(18, 20),
+  `University degree` = c(21, 24),
+  `Bachelor's degree` = c(24, 27),
+  `Above bachelor's degree` = c(28, 32))
+
+# Apply the wage rate ranges to the Avg_hourly_wage_rate column based on Education_level
[…]

### 2024-04-19 scripts/02-data_cleaning.R

@@ -14,20 +14,31 @@ library(arrow)
 #### Clean data ####
 raw_data <- read_csv("data/raw_data/raw_data.csv")
 
-desired_age_groups <- c("15-24 years", "25-54 years", "55 years and over")
+# desired_age_groups <- c("15-24 years", "25-54 years", "55 years and over")
 
 cleaned_data <-
   raw_data |>
   filter(Geography == "Canada",
          `Type of work` == "Both full- and part-time",
          Wages == "Average hourly wage rate",
+         `Education level` != "Some high school",
+         `Education level` != "Some post-secondary",
          `Education level` != "Total, all education levels",
          `Education level` != "PSE  (5,6,7,8,9))",
          `Education level` != "No PSE  (0,1,2,3,4)",
          YEAR >= 2000) |>
-  filter(`Age group` %in% desired_age_groups) |>
+  filter(`Age group` == "25-54 years") |>
   select(YEAR, `Education level`, `Age group`, `Both Sexes`) |>
   rename(Year = YEAR, `Avg hourly wage rate` = `Both Sexes`)
+
+education_levels_ordered <- c("0 - 8  years", "High school graduate", 
+                              "Post-secondary certificate or diploma", "Trade certificate or diploma", 
+                              "Community college, CEGEP", "University certificate below bachelors degree", 
+                              "University degree", "Bachelor's degree", "Above bachelor's degree")
+
+cleaned_data <- cleaned_data %>%
+  mutate(`Education level` = factor(`Education level`, levels = education_levels_ordered)) %>%
+  mutate(Education_numeric = as.numeric(`Education level`))
 
 #### Save data ####
 write_csv(cleaned_data, "data/analysis_data/analysis_data.csv")

### 2024-04-19 scripts/03-test_data.R

@@ -13,3 +13,35 @@ library(tidyverse)
 
 
 #### Test data ####
+# Load dataset
+analysis_data <- read.csv("data/analysis_data/analysis_data.csv", check.names = FALSE)
+
+# Test 1: Check for missing values in simulated_data
+test1_result <- !any(is.na(analysis_data))
+print(paste("Test 1 Passed:", test1_result))
+
+# Test 2: Check for duplicated rows in simulated_data
+test2_result <- !any(duplicated(analysis_data))
+print(paste("Test 2 Passed:", test2_result))
+
+# Test 3: Check that there are exactly 9 unique education levels
+expected_levels <- c("0 - 8  years", "High school graduate", 
+                     "Post-secondary certificate or diploma", "Trade certificate or diploma", 
+                     "Community college, CEGEP", "University certificate below bachelors degree", 
+                     "University degree", "Bachelor's degree", "Above bachelor's degree")
+
+actual_levels <- unique(analysis_data$`Education level`)
+test3_result <- length(actual_levels) == 9 && all(actual_levels %in% expected_levels)
+print(paste("Test 3 Passed:", test3_result))
+
+# Test 4: Check that Education_numeric matches the levels of Education_level
+test4_result <- all(analysis_data$Education_numeric == as.integer(factor(analysis_data$Education_level, levels = education_levels)))
+print(paste("Test 4 Passed:", test4_result))
+
+# Test 5: Check that all years are within the specified range (2000-2019)
+test5_result <- all(analysis_data$Year >= 2000 & analysis_data$Year <= 2019)
+print(paste("Test 5 Passed:", test5_result))
+
+# Test 6: Check that all age groups are "25-54 years"
+test6_result <- all(analysis_data$`Age group` == "25-54 years")
+print(paste("Test 6 Passed:", test6_result))

### 2024-04-19 scripts/04-model.R

@@ -12,14 +12,32 @@
 library(tidyverse)
 library(arrow)
 library(rstanarm)
+library(modelsummary)
+
 
 #### Read data ####
-analysis_data <- read_parquet(file = "data/analysis_data/analysis_data.parquet")
+analysis_data <- read.csv(file = "data/analysis_data/analysis_data.csv")
+
+plot1 = analysis_data |>
+  ggplot(aes(x = Education_numeric, y = Avg.hourly.wage.rate)) +
+  geom_point(alpha = 0.5) +
+  theme_classic() + 
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+
+plot1 +
+  geom_smooth(
+    method = "lm",
+    se = TRUE,
+    color = "black",
+    linetype = "dashed",
+    formula = "y ~ x"
+  )
+
 
 ### Model data ####
 first_model <-
   stan_glm(
-    formula = `Avg hourly wage rate` ~ `Education level` + `Age group`,
+    formula = Avg.hourly.wage.rate ~ Education_numeric,
     data = analysis_data,
     family = gaussian(),
     prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
@@ -27,6 +45,12 @@ first_model <-
     prior_aux = exponential(rate = 1, autoscale = TRUE),
     seed = 853
   )
+
+modelsummary(
+  list(
+    "Auto-scaling priors" = first_model
+  )
+)
 
 
 #### Save model ####

### 2024-04-19 paper/paper.qmd
update model and paper

@@
 library(tinytex)
+library(modelsummary)
 ```
@@
 
-@sec-result presents the core findings from our analysis of US birth rates, specifically focusing on the changes observed over time, across different age groups of young women, and by state.
+@sec-result explores the relationship between education and average hourly wages in Canada.
+Utilizing a dataset that captures 20 years of wage data across varying educational levels, we apply a linear modeling approach to uncover patterns and draw conclusions about the economic value of educational advancement.
+Below, we present the results of our model and its implications.
 
-In conclusion, our findings provide a complete perspective of dropping birth rates in the United States, demonstrating both temporal and demographic tendencies.
-The upcoming @sec-discussion will look at the various variables and biases that may be driving these trends.
+```{r}
+#| label: tbl-summary-model
+#| tbl-cap: Model Summary
+#| echo: false
+#| warning: false
+
+first_model <- readRDS(file = here::here("models/first_model.rds"))
+
+modelsummary(
+  list(
+    "Linear Model" = first_model
+  )
+)
+```
+
+The results obtained from our linear model, which predicts the average hourly wage rate based on the numeric representation of education levels, show a clear and positive relationship between educational attainment and wage rate.
+@tbl-summary-model indicates that the intercept is estimated at 16.143, suggesting that when the education level is at the baseline (0 - not accounted for in the numeric scale), the average hourly wage rate would be approximately 16.14.
+Importantly, the coefficient for Education_numeric is positive (1.672), supporting the hypothesis that higher educational levels correlate with higher wage rates.
+The model used 180 observations for the analysis, with an R-squared value of 0.610, which means that approximately 61% of the variance in the average hourly wage rate is explained by the model.
+The adjusted R-squared, which accounts for the number of predictors in the model, is very close to the R-squared, indicating a good fit without overfitting.
+Notably, the RMSE (Root Mean Square Error) of 3.43 reflects the standard deviation of the prediction errors, which measures how spread out these errors are around the true regression line.
+
+```{r fig.width=7, fig.height=4}
+#| label: fig-visualize-model
+#| fig-cap: Model Visualization
+#| echo: false
+#| warning: false
+
+analysis_data <- read.csv(file = here::here("data/analysis_data/analysis_data.csv"))
+
+plot1 <- analysis_data |>
+  ggplot(aes(x = Education_numeric, y = Avg.hourly.wage.rate, colour = as.factor(Year))) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(
+    method = "lm",
+    se = TRUE,
+    color = "black",
+    linetype = "dashed",
+    formula = y ~ x
+  ) +
+  scale_colour_viridis_d() +  # Colorblind-friendly palette
+  labs(
+    x = "Education Level",  # Rename x-axis
+    y = "AVG Hourly Wage Rate",  # Rename y-axis
+    colour = "Year"  # Rename legend title
+  ) +
+  theme_classic() +
+  theme(
+    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
+    axis.title.x = element_text(size = 12),  # Adjust x-axis title size if needed
+    axis.title.y = element_text(size = 12),  # Adjust y-axis title size if needed
+    legend.position = "right",
+    legend.title = element_text(hjust = 0.5)
+  ) +
+  guides(colour = guide_legend(title = "Year", ncol = 2))
+# Display the plot
+print(plot1)
+```
+@fig-visualize-model illustrates the fitted linear regression line across the spread of average hourly wage rates for different education levels over the years 2000 to 2019.
+Each point in the scatter plot corresponds to the average hourly wage rate for a given education level, differentiated by color to represent each year.
+The dashed line across the plot signifies the best-fit linear regression line, which inclines upward as the education level increases.
+The diversity of colors along the regression line also reveals the variation in wage rates across different years, although the increasing trend is consistent throughout.
+
+In conclusion, the results presented align with the initial hypothesis, showing that as education level increases, there is a corresponding increase in the average hourly wage rate.
+This validates the underlying assumption that more advanced education tends to lead to better-paying job opportunities.
+The analysis effectively captures the essence of the data, allowing for a meaningful interpretation that leads us into the discussion.
+In the next @sec-discussion, we will delve into the implications of these findings, examining their relevance to current labor market theories and the potential for informing education and economic policy decisions.
 
@@
 
-The findings of this study offer a layered perspective on the decline in birth rates among young people in the United States, particularly in the context of the Great Recession.
-Our analysis provides an understanding of the multifaceted influences on demographic trends.
+In this paper, we investigated the relationship between educational attainment and average hourly wage rates from 2000 to 2019.
+By deploying a linear Bayesian model, we have been able to highlight the correlation between higher educational attainment and increased wage rates, aligning with our original hypothesis.
 
@@
 
-In our replication of @citeoriginal, we reaffirmed the central finding that the Great Recession has had a lasting impact on birth rates among young people in the United States.
-Our analysis of Figures 1 and 2 showed a pronounced decline in birth rates across all age groups, with the steepest decrease observed among those aged 15-29.
-This trend extended beyond the economic recovery, suggesting that the Great Recession may have fundamentally altered the family planning trajectory of a generation.
+Our findings reinforce the perception of education as an investment yielding tangible financial returns.
+The analysis showcases a clear, positive trajectory in wage rates corresponding with ascending educational levels.
+This outcome underlines the vital role of education in driving economic growth, suggesting that educational enhancements could be pivotal in strategies aimed at combating poverty and fostering economic resilience.
 
-## Economic Impact Insights
+Further examination reveals the potential of education as a tool for social mobility.
+The data indicates that as individuals climb the educational ladder, they unlock higher earnings, which could lead to a more equitable distribution of income.
+These insights could serve as a valuable asset for policymakers striving for educational equity and highlight the importance of making higher education accessible to all strata of society.
 
-The dramatic drop in birth rates among young Americans during and after the Great Recession demonstrates the profound impact of economic hardship on reproductive decisions.
-This decline in birth rates is not simply coincidental with the date of the recession, but rather indicates a deeper, more systemic influence of economic insecurity on personal life choices, notably the decision to establish or expand a family.
-The persistence of this pattern, even in the years after economic recovery, suggests that the effects of the recession went beyond immediate financial hardship, influencing long-term views of financial security and stability.
-This is further evidenced by the lack of rebound in birth rates post-recession, which might have been anticipated if the decline were solely due to immediate economic pressures.
+## Insights on Education and Wages
 
-The economic model of fertility provides a framework for understanding these patterns by taking into account the cost of pregnancy and raising as well as the opportunity costs associated with parental time and resources.
-The recession likely increased these expenditures and opportunity costs, making the decision to have children more difficult.
-Unemployment and job uncertainty may have caused a reevaluation of the feasibility of having children, resulting in the postponement or avoidance of childbirth.
-Furthermore, the economic downturn may have shifted young adults' expectations and desires for financial security, impacting their family planning decisions.
+The positive correlation observed in this study goes beyond confirming a long-suspected link between education and wages; it also points to the complex interplay of various wage determinants.
+Despite the clear trend, the dispersion of wage data around the regression line suggests multiple underlying factors at play.
+This finding suggests a need for a comprehensive approach to wage enhancement, integrating elements such as vocational training, relocation incentives, and specialized educational programs.
 
-## Societal and Technological Influences
+Expanding on the above, it is evident that education serves as one of many pillars supporting an individual's earning capacity.
+While education remains a significant factor, the disparity between the predicted and observed wages implies other variables are also influential.
+Recognizing this complexity is crucial for a complete understanding of how education impacts wage rates and the formulation of holistic policies that address the multifaceted nature of the job market.
 
-Apart from economic factors, the decline in birth rates among young people also reflects broader societal and technological shifts.
-The period following the Great Recession coincided with significant changes in social norms, increased educational and career opportunities for women, and advancements in reproductive technology.
-These factors collectively have empowered individuals, especially women, to make more independent decisions regarding their reproductive health and family planning.
+## Limitations
 
-As women's educational attainment and labor force involvement have increased, so has the opportunity cost of childbirth, potentially contributing to the drop in birth rates.
-Technological developments have given women greater control over their fertility, allowing for more intentional planning around childbirth.
-Social changes, such as delayed marriage and a growing acceptance of childlessness or smaller family sizes as realistic lifestyle options, further compound these tendencies.
+The primary limitation of this study is its relatively small sample size, with only 180 observations.
+This constraint may impact the robustness and generalizability of the findings.
+Additionally, the focus on the age group of 25-54 years might not encapsulate the full spectrum of the workforce, especially considering the varied career stages and life circumstances of younger and older workers.
+Such a limited scope might overlook the nuanced effects of education on wages across different life stages.
 
-The interaction of these economic, social, and technical elements has altered the landscape of family planning for young Americans.
-While the economic model of fertility provides a framework for understanding these developments, including sociological and technical effects provides a more complete view of the dynamics affecting current birth rate patterns.
+## Future Research Directions
 
-## Weaknesses and Future Research Directions
+Future research should aim to disaggregate the age groups to understand the differential impact of education on wages across a more granular age spectrum.
+Distinguishing between early-career individuals, mid-career professionals, and late-career individuals could yield more nuanced insights into the education-wage dynamic at different life stages.
+Additionally, expanding the sample size and including more varied age groups would enhance the representativeness and depth of the analysis.
 
-One limitation of our study is the potential for unobserved variables that could affect birth rates, such as cultural shifts and changes in social norms, which were not fully captured in the data.
-Additionally, the original study did not account for the influence of the Affordable Care Act and its potential impact on family planning decisions, an area our study also does not explore.
-
-Future research should aim to disentangle these complex relationships further, perhaps through longitudinal studies or by incorporating more detailed data on individual socioeconomic status.
-Understanding these dynamics is crucial for developing policies that support young people in their family planning decisions during and after economic downturns.
+Incorporating a more comprehensive range of variables and exploring more complex, perhaps non-linear models, would allow for a better understanding of the multifarious factors influencing wages.
+Such an enriched analysis could serve as the foundation for more effective, tailored economic policies that address both the direct and indirect ways education can influence economic outcomes.
 

### 2024-04-19 scripts/04-model.R

@@ -18,21 +18,6 @@ library(modelsummary)
 #### Read data ####
 analysis_data <- read.csv(file = "data/analysis_data/analysis_data.csv")
 
-plot1 = analysis_data |>
-  ggplot(aes(x = Education_numeric, y = Avg.hourly.wage.rate)) +
-  geom_point(alpha = 0.5) +
-  theme_classic() + 
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
-
-plot1 +
-  geom_smooth(
-    method = "lm",
-    se = TRUE,
-    color = "black",
-    linetype = "dashed",
-    formula = "y ~ x"
-  )
-
 
 ### Model data ####
 first_model <-
@@ -44,12 +29,6 @@ first_model <-
     prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_aux = exponential(rate = 1, autoscale = TRUE),
     seed = 853
-  )
-
-modelsummary(
-  list(
-    "Auto-scaling priors" = first_model
-  )
 )
 
 

### 2024-04-19 paper/paper.qmd
finalize paper, update readme, update reference, fix grammar

@@
 
-# install.packages("tidyverse")
-# install.packages("dplyr")
-# install.packages("here")
-# install.packages("mapdata")
-# install.packages("ggplot2")
-# install.packages("haven")
-# install.packages("kableExtra")
-# install.packages('tinytex')
-# tinytex::install_tinytex()
-# to uninstall TinyTeX, run tinytex::uninstall_tinytex()
-
 library(tidyverse)
@@
 library(here)
-library(mapdata)
 library(ggplot2)
@@
 library(kableExtra)
-library(tinytex)
 library(modelsummary)
+library(rstanarm)
 ```
@@
 
+The structure of the paper is organized as follows: Following @sec-intro, @sec-data presents the data, detailing the data sources, analytical techniques, and the rationale behind the chosen methods.
+@sec-model then covers the specifics of the linear model analysis, laying out the statistical underpinnings that support our investigation.
+After that, @sec-result discusses the results, elaborating on the observed trends and patterns in wage rate data.
+@sec-discussion interprets these findings in light of the current economic and educational context in Canada, exploring potential factors influencing these trends, drawing connections to broader socio-economic issues, and providing suggestions for future research in this area.
+
+## Estimand
+
 In this study, our primary focus is to estimate the effect of education on hourly wages in Canada.
 Our estimand is the incremental average hourly wage rate for individuals within the specified education levels, compared to the next lower education level.
-By quantifying this effect, we aim to capture the economic value of educational attainment and how it translates into wage premiums across various stages of a typical Canadian worker’s career span.
-
-The structure of the paper is organized as follows: Following @sec-intro, @sec-data presents the data, detailing the data sources, analytical techniques, and the rationale behind the chosen methods.
-@sec-model then delves into the specifics of the linear model analysis, laying out the statistical underpinnings that support our investigation.
-After that, @sec-result discusses the results, elaborating on the observed trends and patterns in wage rate data.
-@sec-discussion interprets these findings in light of the current economic and educational context in Canada, exploring potential factors influencing these trends, drawing connections to broader socio-economic issues, and providing suggestions for future research in this area.
+By quantifying this effect, we aim to capture the economic value of educational attainment and how it translates into wage premiums.
 
@@
 
-This study utilizes a dataset sourced from the Open Government Portal of Canada [], specifically designed to track the correlation between educational attainment and hourly wages across various demographic segments of the Canadian workforce.
+This study uses a dataset sourced from the Open Government Portal of Canada [@citedata], specifically designed to track the correlation between educational attainment and hourly wages across various demographic segments of the Canadian workforce.
 The dataset covers the period from the year 2000 onwards, providing a longitudinal view of wage trends in relation to educational background.
@@
 This step was crucial to maintaining the clarity and relevance of the analysis.
-For key operations, please refer to the [@sec-appendix].
+For key operations, please refer to the [Appendix -@sec-data-cleaning].
 
@@
 #| label: tbl-display-clean-data
-#| tbl-cap: First Ten Rows of Wage Rates Spanning from 2000 to 2019 For Cleaned Data
+#| tbl-cap: First Ten Rows of Hourly Wage Across 2000-2019 After Data Processing
 #| echo: false
@@
 #| label: fig-display-clean-data
-#| fig-cap: Visualization of Wage Rate from 2000 to 2019
+#| fig-cap: Average Hourly Wages by Educational Attainment, 2000-2019
 #| echo: false
@@
   ) +
-  theme_minimal(base_size = 10) +  # Reduced base_size for smaller overall text
+  theme_minimal(base_size = 10) + 
   theme(
-    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8),  # Smaller font for x-axis labels
-    axis.text.y = element_text(size = 8),  # Smaller font for y-axis labels
+    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8),
+    axis.text.y = element_text(size = 8),
     legend.position = "right",
-    legend.text = element_text(size = 8)  # Smaller font for legend text
+    legend.text = element_text(size = 8)
   ) +
@@
 ```
-@fig-display-clean-data illustrates the distribution of average hourly wage rates across different educational levels from 2000 to 2019 for Canadian workers aged 25-54.
+@fig-display-clean-data built by `ggplot2` [@citeggplot] illustrates the distribution of average hourly wage rates across different educational levels from 2000 to 2019 for Canadian workers aged 25-54.
 Each dot represents the wage rate for a specific education level in a given year, with color coding to distinguish between educational categories, ranging from "0 – 8 years" to "University degree." The spread and upward trend of the points suggest that individuals with higher educational qualifications tend to have higher average hourly wages, and this trend persists over the 20-year period.
@@
 The 'Average Hourly Wage Rate' is calculated based on actual wages reported by employers, providing an objective measure of income.
-This data is then vetted and standardized by statistical agencies to ensure it reflects accurate and fair representations of wages across different demographics.
+This data is then vetted and standardized by the Government of Ontario to ensure it reflects accurate and fair representations of wages across different demographics.
 
@@
 This data is collected through surveys and employment records, often as part of broader demographic data collection efforts by governmental agencies.
-
-For an in-depth technical breakdown of the statistical analyses and additional visualizations, readers are directed to the appendices, where these aspects are discussed in detail.
-This ensures that the main text remains accessible while still providing resources for more technically inclined readers.
 
@@
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+In our analysis, we utilized a simple linear regression model, a technique well-suited for examining the relationship between a continuous dependent variable and one or more independent variables.
+Given that our dependent variable, `average hourly wage rate`, is continuous, a normal or Gaussian linear regression model is appropriate.
+This model assumes that the distribution of the wage rate, given the level of education, follows a normal distribution.
+This assumption allows for a straightforward interpretation of the parameters and is commonly used in practice.
+The use of a normal distribution for the error term is a standard choice, as it facilitates the estimation of the model parameters using maximum likelihood estimation.
+The simplicity and interpretability of this model, coupled with its appropriateness for the data at hand, make it a good fit for our study's objectives.
+
+Background details and diagnostics are included in [Appendix -@sec-model-details].
 
@@
 $$
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
 
-The model is executed in R[@citeR] using the rstanarm package [@rstanarm].
-Default priors from rstanarm are utilized, set with a mean of zero and a conservative standard deviation, reflecting no strong prior beliefs about the magnitude of the education coefficients.
+The model is executed in R [@citeR] using the `rstanarm` package [@citerstanarm].
+Default priors from `rstanarm` [@citerstanarm] are used, set with a mean of zero and a conservative standard deviation.
 
@@
 
-Given the economic theory and previous empirical findings, we anticipate that higher educational qualifications would correspond to higher hourly wages.
+Given the economic theory and previous findings, we anticipate that higher educational qualifications would correspond to higher hourly wages.
 This is encapsulated in the assumption that $\beta_i$ should have a positive effect on $y_i$.
-Through Bayesian analysis, we aim to capture the degree to which educational attainment can predict wage rates, thus providing empirical evidence to support or refine this hypothesis.
+Through Bayesian analysis, we aim to capture the degree to which educational attainment can predict wage rates, thus providing evidence to support or refine this hypothesis.
 
-Mathematical notation is incorporated using LaTeX syntax, enabling precise formulation of our statistical model. For example, the average effect of education on hourly wage rate is denoted by $\beta$.
+We selected a Gaussian linear regression model to examine the relationship between education level and average hourly wages because our dependent variable, the wage rate, is a continuous data point that typically follows a normal distribution.
+Linear regression is a straightforward method for assessing the impact of one or more independent variables on a dependent variable.
+In this case, it helps us understand how increasing education levels, which we have numerically encoded, relate to wage rates.
+
+Further justification for this model comes from its foundation in the central limit theorem, which suggests that with a large sample size, our wage rate data should be normally distributed.
+Moreover, the expected positive correlation between education and wages aligns with existing economic studies, reinforcing our theoretical framework.
+The simplicity of the model aids in avoiding overfitting, ensuring that the results are generalizable and relevant to a broader population.
+The regression model's assumptions—linearity, independence, and normality of residuals—are all conditions that our dataset reasonably meets, making a Gaussian linear regression not only a convenient choice but also statistically appropriate for our analysis.
 
@@
 #| label: tbl-summary-model
-#| tbl-cap: Model Summary
+#| tbl-cap: Summary of the Linear Regression Model
 #| echo: false
@@
 
-first_model <- readRDS(file = here::here("models/first_model.rds"))
+linear_model <- readRDS(file = here::here("models/linear_model.rds"))
 
@@
   list(
-    "Linear Model" = first_model
+    "Linear Model" = linear_model
   )
@@
 #| label: fig-visualize-model
-#| fig-cap: Model Visualization
+#| fig-cap: Relationship between Hourly Wage Rate and Education Level
 #| echo: false
@@
 The analysis effectively captures the essence of the data, allowing for a meaningful interpretation that leads us into the discussion.
-In the next @sec-discussion, we will delve into the implications of these findings, examining their relevance to current labor market theories and the potential for informing education and economic policy decisions.
-
-\newpage
+In the next @sec-discussion, we will discuss the implications of these findings, examining their relevance to current labor market theories and the potential for informing education and economic policy decisions.
 
@@
 
-Our findings reinforce the perception of education as an investment yielding tangible financial returns.
-The analysis showcases a clear, positive trajectory in wage rates corresponding with ascending educational levels.
-This outcome underlines the vital role of education in driving economic growth, suggesting that educational enhancements could be pivotal in strategies aimed at combating poverty and fostering economic resilience.
-
-Further examination reveals the potential of education as a tool for social mobility.
-The data indicates that as individuals climb the educational ladder, they unlock higher earnings, which could lead to a more equitable distribution of income.
-These insights could serve as a valuable asset for policymakers striving for educational equity and highlight the importance of making higher education accessible to all strata of society.
+Our analysis reveals a clear, positive correlation between education level and average hourly wage rates across the data spanning from 2000 to 2019.
+The consistent upward trend in wage rates with higher education levels across the years underscores the value of educational attainment in the labor market.
+Despite some fluctuations, the general pattern suggests that investments in education are likely to be economically beneficial for individuals in the age group of 25-54 years.
 
@@
 
-The positive correlation observed in this study goes beyond confirming a long-suspected link between education and wages; it also points to the complex interplay of various wage determinants.
-Despite the clear trend, the dispersion of wage data around the regression line suggests multiple underlying factors at play.
-This finding suggests a need for a comprehensive approach to wage enhancement, integrating elements such as vocational training, relocation incentives, and specialized educational programs.
+The results of this study offer clear evidence that educational attainment is closely tied to wage earnings, highlighting the premium that the labor market places on higher qualifications.
+This finding is consistent with current labor market theories that emphasize the importance of human capital—suggesting that investments in education are indeed reflected in higher wages.
+Specifically, our data supports the view that a well-educated workforce is critical for economic growth and competitiveness in today's globalized world, which increasingly values knowledge and specialized skills.
 
-Expanding on the above, it is evident that education serves as one of many pillars supporting an individual's earning capacity.
-While education remains a significant factor, the disparity between the predicted and observed wages implies other variables are also influential.
-Recognizing this complexity is crucial for a complete understanding of how education impacts wage rates and the formulation of holistic policies that address the multifaceted nature of the job market.
+From a policy perspective, these insights could serve as a catalyst for reform, underscoring the need for investments in education.
+Policymakers might use this information to support sectors that are likely to offer greater economic returns for educational investments.
+This could involve bolstering STEM (Science, Technology, Engineering, and Mathematics) education, for example, or providing incentives for pursuing in-demand fields such as healthcare or information technology.
 
-## Limitations
+In conclusion, this study's findings highlight the significance of education as a determinant of economic success and reinforce the importance of informed educational and economic policymaking.
+For policymakers, the imperative is clear: to design and implement educational strategies that not only increase access to education but also ensure that the education provided is relevant and responsive to the current and future needs of the labor market.
+
+## Limitations and Future Research Directions
 
 The primary limitation of this study is its relatively small sample size, with only 180 observations.
-This constraint may impact the robustness and generalizability of the findings.
+This constraint may impact the generalization of the findings.
 Additionally, the focus on the age group of 25-54 years might not encapsulate the full spectrum of the workforce, especially considering the varied career stages and life circumstances of younger and older workers.
-Such a limited scope might overlook the nuanced effects of education on wages across different life stages.
+Such a limited scope might overlook the effects of education on wages across different life stages.
 
-## Future Research Directions
-
-Future research should aim to disaggregate the age groups to understand the differential impact of education on wages across a more granular age spectrum.
-Distinguishing between early-career individuals, mid-career professionals, and late-career individuals could yield more nuanced insights into the education-wage dynamic at different life stages.
+Future research should aim to disaggregate the age groups to understand the differential impact of education on wages.
+Distinguishing between early-career individuals, mid-career professionals, and late-career individuals could yield more insights into the education-wage dynamic at different life stages.
 Additionally, expanding the sample size and including more varied age groups would enhance the representativeness and depth of the analysis.
-
-Incorporating a more comprehensive range of variables and exploring more complex, perhaps non-linear models, would allow for a better understanding of the multifarious factors influencing wages.
-Such an enriched analysis could serve as the foundation for more effective, tailored economic policies that address both the direct and indirect ways education can influence economic outcomes.
 
@@
 
-# Data Manipulation and Cleaning {#sec-appendix}
+# Data Manipulation and Cleaning {#sec-data-cleaning}
 
-Most of the data in our dataset was previously cleaned for the project we are replicating.
-Thus, @fig-trends-in-birth-rate directly used data from `outputs/data/fig_1.csv` without any further cleaning necessary, as it simply displays the birth rate from 1980 to 2020.
-Thus, the cleaned data is directly stored to `outputs/data/fig_1.csv` from `inputs/data/fig_1.csv`.
+During the data cleaning phase, the R packages `tidyverse` [@citetidyverse], `dplyr` [@citedplyr], and `arrow` [@citearrow] were utilized.
+The raw data was imported using `read_csv` from the `tidyverse` [@citetidyverse] package.
+Subsequent operations filtered the dataset to focus on Canadian data, both full-time and part-time.
+Additionally, the focus was placed on records that reported the average hourly wage rate, while entries with ambiguous educational levels such as "Some high school" and aggregate categories like "Total, all education levels" were excluded to maintain data clarity and relevance.
 
-In @fig-trends-in-birth-rate-in-young-people, the dataset was imported from `inputs/data/fig_2.csv` and used `dpylr` [@citedplyr] to select the birth rates of the six age groups (from 15 to 44).
-Furthermore, the selected data was renamed using `tidyverse` [@citetidyverse] and changed into names that represent the content of the data better.
-Moreover, these data is then pivoted using `tidyverse` [@citetidyverse] in order to be graphed properly using `ggplot2` [@citeggplot].
+The analysis was specifically targeted at the demographic group aged between 25 and 54 years, as this range represents a prime working-age population.
+Columns important to the analysis were chosen and renamed for better readability and straightforward reference in subsequent analytical procedures.
+This process included renaming the columns for the year and average hourly wage rate to 'Year' and 'Avg hourly wage rate' respectively.
 
-In @fig-birth-rate-comparison-old-and-young, the first dataset was imported from `inputs/data/fig_3.csv` and the dataset containing American states information was from `mapdata` [@citemapdata].
-The first dataset was first sliced to remove birth rate information regarding the states of Alaska and Hawaii, as these two states can't be properly shown in the dataset from `mapdata` [@citemapdata].
-Then, a temporary data frame is created with tibble from `tidyverse` [@citetidyverse] with the abbreviated state names from `inputs/data/fig_3.csv` and the state full names.
-Moreover, the abbreviated state name in the first dataset is switched with the state full name using left_join from `dpylr` [@citedplyr], which is then selected and renamed using `tidyverse` [@citetidyverse].
-After this, the adjusted dataset with the state full names and birth rate is merged with the second dataset that contains the American states information for further graphing with `haven` [@citehaven].
+Education levels were organized in a logical order ranging from "0 - 8 years" of education to "Above bachelor's degree" using the `factor` function to convert them into a categorical variable with a specified level order.
+This categorization was further enhanced by creating an accompanying numeric variable that mapped these ordered education levels to integers, thus facilitating quantitative analysis.
 
-In all three figures, the library `here` [@citehere] was used to ensure that the file path should be accessible in all directories.
+The final step involved saving the cleaned and structured data.
+This was done using the `write_csv` function to generate a CSV file for broad compatibility and the `write_parquet` function from the `arrow` [@citearrow] package for a more compressed and efficient file format, both of which were stored in the `data/analysis_data` directory.
 
-The transformation from raw to analysis-ready data involved substantial cleaning, including filtering by geography (Canada), type of work (both full- and part-time), and specific wage type (average hourly rate).
-Moreover, we excluded broad or ambiguous categories like 'Some high school' and 'Total, all education levels' to ensure clarity in our analysis.
-The resulting 'Education_numeric' variable was constructed using a sorted factorization process, mapping each education level to a corresponding numeric value.
+In all figures and tables, the library `here` [@citehere] was used to ensure that the file path should be accessible in all directories.
 
-In presenting our findings, we will incorporate graphs that visualize the relationship between education and wage rates, providing an intuitive understanding of the data's characteristics.
-Tables summarizing the observations, along with discussions on the graphs, will be included to elucidate key trends and observations.
-Summary statistics and any notable relationships between the variables will also be featured, offering a detailed overview of the dataset.
-Should the volume of this information warrant it additional detail, these elements may be relegated to appendices to maintain the accessibility of the data section.
+# Model details {#sec-model-details}
+
+## Posterior predictive check
+
+In @fig-ppcheckandposteriorvsprior-1, we implement a posterior predictive check, which displays the overlap between the observed data (denoted by $y$) and the replicated data generated from the model (denoted by $y_{rep}$).
+The density of the replicated data largely coincides with the observed data, suggesting that our model does a reasonable job of capturing the underlying distribution of the observed average hourly wage rates.
+This alignment is crucial as it indicates that the model's predictions are consistent with the real-world data we aim to understand and that the model can be a reliable tool for further inference.
+
+In @fig-ppcheckandposteriorvsprior-2, we compare the posterior distributions of our parameters with the priors.
+This comparison reveals how the observed data have updated our beliefs about the parameters.
+We see that the posterior distributions for the intercept and the education_numeric coefficient are more concentrated and shifted from the priors, indicating that the data provided substantial information to update our initial assumptions.
+The sigma parameter's posterior, which represents the standard deviation of the residuals, is also refined from its prior, showing that the data have informed us about the variability in wages that is not explained by education level alone.
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-ppcheckandposteriorvsprior
+#| layout-ncol: 2
+#| fig-cap: "Examining how the model fits, and is affected by, the data"
+#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
+
+pp_check(linear_model) +
+  theme_classic() +
+  theme(legend.position = "bottom")
+
+posterior_vs_prior(linear_model) +
+  theme_minimal() +
+  scale_color_brewer(palette = "Set1") +
+  theme(legend.position = "bottom") +
+  coord_flip()
+```
+
+## Diagnostics
+
+```{r}
+#| echo: false
+#| eval: true
+#| message: false
+#| warning: false
+#| label: fig-stanareyouokay
+#| fig-cap: "Checking the convergence of the MCMC algorithm"
+#| fig-subcap: ["Trace plot", "Rhat plot"]
+#| layout-ncol: 2
+
+plot(linear_model, "trace")
+
+plot(linear_model, "rhat")
+```
+
+In @fig-stanareyouokay-1, the trace plot demonstrates the sampling paths of the Markov chains for each parameter in the model.
+The trace for the intercept, education_numeric coefficient, and sigma parameter all display a ‘hairy caterpillar’ shape, indicating good mixing and convergence across multiple chains.
+This suggests that the model's parameter estimates are stable and reliable, as evidenced by the overlapping and intertwined paths, which is indicative of a well-fitting model.
+
+@fig-stanareyouokay-2, the Rhat plot, presents the R-hat statistics for each parameter, which is a measure of convergence.
+We see that all values at or very close to 1.00.
+This is a desirable outcome, implying that further sampling would likely yield little additional information and that the posterior estimates can be considered reliable for inference.
 

### 2024-04-19 scripts/00-install_packages.R

@@ -4,8 +4,8 @@
 # Date: 31 March 2024
 # Contact: leoy.liu@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: -
+# Any other information needed? -
 
 
 #### Install packages ####
@@ -13,3 +13,7 @@ install.packages("tidyverse")
 install.packages("arrow")
 install.packages("dplyr")
 install.packages("rstanarm")
+install.packages("here")
+install.packages("ggplot2")
+install.packages("kableExtra")
+install.packages("modelsummary")

### 2024-04-19 scripts/01-simulate_data.R

@@ -4,20 +4,16 @@
 # Email: leoy.liu@mail.utoronto.ca
 # Date: 31 March 2024
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: -
+# Any other information needed? -
 
 
 #### Workspace setup ####
 library(tidyverse)
 library(dplyr)
 
-#### Simulate data ####
-# In this simulation, we focus on two different age categories for birth rate data: younger and older groups.
-# The simulation starts with birth rate of 6 age groups.
-# The simulated data is derived from a normal distribution instead of a poisson distribution since birth rate can be more than just integers.
-# By applying the abs() function, we guarantee the recording of only positive birth rates for the older age group.
 
+#### Simulate data ####
 set.seed(853)
 
 # Set parameters for simulation

### 2024-04-19 scripts/02-data_cleaning.R

@@ -4,17 +4,17 @@
 # Date: 31 March 2024
 # Contact: leoy.liu@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: raw_data.csv
+# Any other information needed? -
+
 
 #### Workspace setup ####
 library(tidyverse)
 library(arrow)
 
+
 #### Clean data ####
 raw_data <- read_csv("data/raw_data/raw_data.csv")
-
-# desired_age_groups <- c("15-24 years", "25-54 years", "55 years and over")
 
 cleaned_data <-
   raw_data |>
@@ -39,6 +39,7 @@ education_levels_ordered <- c("0 - 8  years", "High school graduate",
 cleaned_data <- cleaned_data %>%
   mutate(`Education level` = factor(`Education level`, levels = education_levels_ordered)) %>%
   mutate(Education_numeric = as.numeric(`Education level`))
+
 
 #### Save data ####
 write_csv(cleaned_data, "data/analysis_data/analysis_data.csv")

### 2024-04-19 scripts/03-test_data.R

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Tests
+# Purpose: Tests for dataset
 # Author: Yuanyi (Leo) Liu
 # Date: 31 March 2024
 # Contact: leoy.liu@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 02-data_cleaning.R
+# Any other information needed? -
 
 
 #### Workspace setup ####

### 2024-04-19 scripts/04-model.R

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Models
+# Purpose: Models the dataset with simple linear regression
 # Author: Yuanyi (Leo) Liu
 # Date: 31 March 2024
 # Contact: leoy.liu@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 02-data_cleaning.R
+# Any other information needed? -
 
 
 #### Workspace setup ####
@@ -29,11 +29,11 @@ first_model <-
     prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_aux = exponential(rate = 1, autoscale = TRUE),
     seed = 853
-)
+  )
 
 
 #### Save model ####
 saveRDS(
   first_model,
-  file = "models/first_model.rds"
+  file = "models/linear_model.rds"
 )

### 2024-04-19 paper/paper.qmd
fix grammar

@@
 library(ggplot2)
-library(haven)
 library(kableExtra)
@@
 
-The data was processed and cleaned using R[@citeR], a powerful statistical programming language.
+The data was processed and cleaned using `R` [@citeR], a powerful statistical programming language.
 The selection of this dataset was motivated by its direct relevance to the research question and its regular updates, ensuring that the data remains relevant for observing current trends.
@@
 This step was crucial to maintaining the clarity and relevance of the analysis.
-For key operations, please refer to the [Appendix -@sec-data-cleaning].
+For key operations, please refer to [Appendix -@sec-data-cleaning].
 
@@
 
-To better understand data, key variables extracted for this study include the year of data collection, education level, age group, and the average hourly wage rate.
+To better understand the data, key variables extracted for this study include the year of data collection, education level, age group, and the average hourly wage rate.
 Notably, we have streamlined the 'Education level' from a broad range of categories into a numeric variable that aligns with ascending educational attainment, facilitating a quantitative analysis of its impact on wages.
@@
 ```
-@fig-display-clean-data built by `ggplot2` [@citeggplot] illustrates the distribution of average hourly wage rates across different educational levels from 2000 to 2019 for Canadian workers aged 25-54.
+@fig-display-clean-data, built by `ggplot2` [@citeggplot], illustrates the distribution of average hourly wage rates across different educational levels from 2000 to 2019 for Canadian workers aged 25-54.
 Each dot represents the wage rate for a specific education level in a given year, with color coding to distinguish between educational categories, ranging from "0 – 8 years" to "University degree." The spread and upward trend of the points suggest that individuals with higher educational qualifications tend to have higher average hourly wages, and this trend persists over the 20-year period.
@@
 
-Correlations between the variables will be examined at @sec-model, providing insights into how education levels are associated with hourly wages.
+Correlations between the variables will be examined in @sec-model, providing insights into how education levels are associated with hourly wages.
 By understanding these relationships, we can better infer the potential impact of education on earnings within the Canadian economy.
@@
 
-The model is executed in R [@citeR] using the `rstanarm` package [@citerstanarm].
+The model is executed in `R` [@citeR] using the `rstanarm` package [@citerstanarm].
 Default priors from `rstanarm` [@citerstanarm] are used, set with a mean of zero and a conservative standard deviation.
